\documentclass[twocolumn]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage{url}
\usepackage{cite}
\usepackage{braket}  % Dirac notation

\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

\title{\textbf{EQ-GAN: Investigating Quantum Entanglement as a Style Injection\\
Mechanism in Hybrid Generative Adversarial Networks}}

\author{
  Felipe Mahlow\\
  \small \texttt{felipe.mahlow@gmail.com}
}

\date{\today}

\begin{document}

\maketitle

% ============================================================
\begin{abstract}
Quantum Generative Adversarial Networks (QGANs) have emerged as a promising
framework for quantum-enhanced generative modeling, yet the specific role of
quantum entanglement as a \emph{computational resource} for generative control
remains poorly understood. We propose \textbf{EQ-GAN}, a hybrid QGAN designed
specifically to test the hypothesis that quantum entanglement can serve as an
effective style injection mechanism for conditional image generation.
EQ-GAN introduces two dedicated quantum registers—a content register
(6 qubits, angle-encoded from latent noise) and a style register (2 qubits,
IQP feature-mapped from class labels)—connected by a structured CNOT
entanglement core that creates cross-register correlations before variational
layers are applied. We conduct a systematic ablation study across FashionMNIST,
MNIST, and BreastMNIST, comparing EQ-GAN against a no-entanglement variant
(independent registers), an unconditional variant (no style register), and a
Patch QGAN baseline. Contrary to our initial hypothesis, the no-entanglement
variant consistently outperforms EQ-GAN with a FID gap of 18.9 points on
FashionMNIST (173.7 vs.\ 192.6), suggesting that the CNOT-based entanglement
core impedes rather than facilitates generative quality in the NISQ simulation
regime. We characterize this finding via circuit expressibility analysis
(KL divergence from the Haar measure) and von~Neumann entanglement entropy
measurements, and discuss implications for entanglement as a computational
resource in near-term quantum generative models.
\end{abstract}

\textbf{Keywords:} quantum GAN, entanglement, hybrid quantum-classical,
conditional image generation, NISQ, parametric quantum circuits, expressibility

% ============================================================
\section{Introduction}
\label{sec:intro}

Generative Adversarial Networks (GANs) \cite{goodfellow2014generative} have
transformed the landscape of generative modeling, enabling the synthesis of
high-fidelity images through an adversarial training framework between a
generator $G$ and discriminator $D$. Classical advances—including the
Wasserstein formulation \cite{arjovsky2017wasserstein}, style-based
architectures \cite{karras2019style}, and self-attention mechanisms
\cite{zhang2019self}—have progressively improved image quality, diversity,
and controllability.

Simultaneously, the emergence of near-term quantum processors has motivated
the study of quantum-enhanced generative models. The theoretical appeal is
rooted in the exponential dimensionality of Hilbert space: an $n$-qubit
system spans a $2^n$-dimensional state space, potentially offering a
representational advantage for capturing complex distributions
\cite{biamonte2017quantum}. Quantum GANs (QGANs) were first proposed by
Lloyd and Weedbrook \cite{lloyd2018quantum} and Dallaire-Demers and Killoran
\cite{dallaire2018quantum}, establishing the theoretical foundations for
adversarial training with quantum generators. Subsequent hybrid architectures
have demonstrated practical generation of low-dimensional data on near-term
hardware \cite{huang2021experimental, tsang2023mosaiq, rudolph2024latent}.

However, a critical gap persists in existing QGAN architectures: \emph{quantum
entanglement is not exploited as an explicit mechanism for generative control}.
Most hybrid QGANs employ PQCs as generic function approximators, using
entanglement implicitly (through CNOT-rich ansatze) without differentiating
between the roles of content generation and style conditioning in quantum
space. This stands in contrast to classical advances such as StyleGAN
\cite{karras2019style}, where disentangled style codes enable precise control
over generative attributes through adaptive instance normalization.

We propose \textbf{EQ-GAN} (Entanglement-enhanced Quantum GAN) as a testbed
architecture to empirically investigate this question. EQ-GAN introduces a
dedicated \emph{entanglement core}—a structured CNOT ladder connecting a style
qubit register (encoding class information via IQP feature maps) to a content
qubit register (encoding latent noise via angle encoding). Unlike prior hybrid
QGANs where entanglement arises implicitly from circuit topology, EQ-GAN
makes entanglement between content and style registers a first-class design
choice, enabling a controlled ablation of its role.

\paragraph{Contributions.} This paper makes four contributions:
\begin{enumerate}
  \item \textbf{EQ-GAN architecture}: a hybrid QGAN with 8-qubit circuit
    (6 content, 2 style) with a structured CNOT entanglement core, designed
    as a controlled testbed for studying entanglement-as-style-injection.
  \item \textbf{Negative result}: through systematic ablation, we demonstrate
    that the entanglement core does \emph{not} improve—and consistently
    impedes—conditional generation quality. This is a practically important
    finding for QGAN design in the NISQ regime.
  \item \textbf{Quantum characterization}: expressibility analysis (KL from
    Haar measure) and von~Neumann entanglement entropy measurements that
    provide a mechanistic explanation for the observed degradation.
  \item \textbf{Reproducible benchmark}: evaluation on FashionMNIST, MNIST, and
    BreastMNIST with FID, IS, and Precision/Recall metrics, with full code release.
\end{enumerate}

The remainder of the paper is organized as follows. Section~\ref{sec:related}
reviews related work. Section~\ref{sec:method} describes the EQ-GAN
architecture. Section~\ref{sec:theory} presents the theoretical
characterization. Section~\ref{sec:experiments} details the experimental
setup. Section~\ref{sec:results} presents and discusses results.
Section~\ref{sec:conclusion} concludes with limitations and future directions.

% ============================================================
\section{Related Work}
\label{sec:related}

\subsection{Classical GANs and Style Control}

The original GAN \cite{goodfellow2014generative} frames generative modeling
as a minimax game: $\min_G \max_D \mathbb{E}[\log D(x)] +
\mathbb{E}[\log(1-D(G(z)))]$. Conditional GANs \cite{mirza2014conditional}
introduced class-conditional generation by concatenating label information to
both generator and discriminator inputs. The Wasserstein GAN with gradient
penalty (WGAN-GP) \cite{gulrajani2017improved} replaced the Jensen-Shannon
divergence with the Earth Mover's distance, providing stable training dynamics.

Style-based generation \cite{karras2019style, karras2020analyzing} introduced
a fundamentally different conditioning paradigm: instead of concatenating class
labels, a learned mapping network transforms latent codes into \emph{style
parameters} that are injected at each layer via adaptive instance normalization.
This disentanglement of style and content has been highly influential—our
entanglement core is conceptually motivated by this paradigm, transposing it
into quantum space via CNOT-based cross-register correlations.

\subsection{Quantum Generative Adversarial Networks}

Lloyd and Weedbrook \cite{lloyd2018quantum} first proposed a fully quantum GAN
where both generator and discriminator are quantum channels, establishing
theoretical bounds on training dynamics and convergence. Dallaire-Demers and
Killoran \cite{dallaire2018quantum} introduced a framework for training QGANs
on quantum hardware using a quantum cost function based on the diamond norm.

Hybrid architectures—combining quantum generators with classical
discriminators—have dominated practical QGAN research. Huang et al.\
\cite{huang2021experimental} demonstrated a patch-based hybrid QGAN on
superconducting hardware, achieving generation of $8\times 8$ handwritten
digits by assembling image patches from separate PQC outputs. The MosaIQ
architecture \cite{tsang2023mosaiq} extended this with a mosaic strategy,
generating overlapping patches and aggregating them to reduce boundary
artifacts.

Latent QGANs \cite{rudolph2024latent} operate in a compressed latent space
produced by a classical autoencoder, reducing the dimensionality challenge
while enabling the quantum circuit to focus on the semantic structure of the
data. Recent work has explored quantum Wasserstein distances
\cite{kiani2022quantum}, quantum normalizing flows
\cite{johansen2024quantum}, and barren plateau mitigation strategies
\cite{mcclean2018barren, zhang2024trainability} relevant to QGAN training.

Our work differs from existing hybrid QGANs in that we explicitly differentiate
the roles of content and style registers and use entanglement as a structured
conditioning mechanism rather than incidental circuit topology.

\subsection{Entanglement as a Computational Resource}

Entanglement entropy has been studied as a measure of quantum circuit
expressibility \cite{sim2019expressibility}. Abbas et al.\
\cite{abbas2021power} demonstrated that quantum models can achieve higher
effective dimension than comparable classical models. Bipartite entanglement
in PQCs has been linked to trainability \cite{cerezo2021cost} and generalization
\cite{caro2022generalization} in quantum machine learning.

In the context of generative models, entanglement enables the generation of
states with non-classical correlations \cite{lloyd2018quantum}. Our work
provides empirical evidence connecting entanglement entropy to generative
quality in a hybrid setting, operationalizing entanglement as a style
injection mechanism.

% ============================================================
\section{EQ-GAN Architecture}
\label{sec:method}

\subsection{Overview}

EQ-GAN is a conditional hybrid QGAN with five components: (1) a
classical style encoder mapping class labels to quantum style inputs,
(2) a classical noise encoder mapping latent noise to quantum content inputs,
(3) a quantum circuit implementing content encoding, IQP-based style encoding,
an entanglement core, and variational layers,
(4) a classical convolutional decoder mapping quantum measurements to images, and
(5) a conditional CNN discriminator. The full architecture is shown in
Figure~\ref{fig:architecture}.

\subsection{Classical Encoders}

\paragraph{Noise encoder.}
Given Gaussian noise $\mathbf{z} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_{64})$,
the noise encoder $E_\text{noise}: \mathbb{R}^{64} \rightarrow [-\pi, \pi]^{N_C}$
is a two-layer MLP that produces $N_C = 6$ angle parameters $\boldsymbol{\theta}_n$
scaled to $[-\pi, \pi]$ via $\tanh(\cdot) \times \pi$:
\begin{equation}
  \boldsymbol{\theta}_n = \pi \cdot \tanh(W_2 \cdot \text{LeakyReLU}(W_1 \mathbf{z} + b_1) + b_2).
\end{equation}

\paragraph{Style encoder.}
Given a one-hot class label $\mathbf{c} \in \{0,1\}^{N_\text{cls}}$, the style
encoder $E_\text{style}: \mathbb{R}^{N_\text{cls}} \rightarrow [-\pi,\pi]^{N_S}$
produces $N_S = 2$ style parameters $\boldsymbol{\phi}_s$ via a similar two-layer
MLP.

\subsection{Quantum Circuit}
\label{sec:circuit}

The quantum circuit operates on $N_Q = 8$ qubits partitioned into a
\emph{content register} $\mathcal{C} = \{q_0, \ldots, q_5\}$ and a
\emph{style register} $\mathcal{S} = \{q_6, q_7\}$.

\paragraph{Content encoding.} Angle encoding applies individual $R_Y$ rotations
to the content register:
\begin{equation}
  U_n(\boldsymbol{\theta}_n) = \bigotimes_{i=0}^{5} R_Y(\theta_{n,i}).
\end{equation}

\paragraph{Style encoding (IQP feature map).}
The style register uses an IQP-inspired feature map
\cite{havlicek2019supervised}, which is known to produce states that are
classically hard to simulate for random inputs:
\begin{equation}
  U_s(\boldsymbol{\phi}_s) =
    \text{IsingZZ}(\phi_{s,0} \cdot \phi_{s,1}, \{q_6, q_7\}) \cdot
    \bigotimes_{j=6}^{7} [R_Z(\phi_{s,j-6}) \cdot H_{q_j}].
\end{equation}

\paragraph{Entanglement core.}
The entanglement core $U_E$ is the key architectural contribution. It consists
of a CNOT ladder connecting the style register to the content register,
followed by intra-style entanglement:
\begin{equation}
  U_E = \text{CNOT}_{q_7 \to q_6} \cdot \text{CNOT}_{q_6 \to q_0} \cdot \text{CNOT}_{q_7 \to q_1},
  \label{eq:entcore}
\end{equation}
where CNOT$_{a \to b}$ denotes a CNOT gate with control qubit $a$ and target
qubit $b$. This structure creates quantum correlations between the style and
content registers, entangling the class information encoded in $\mathcal{S}$
with the content generation subspace $\mathcal{C}$.

\paragraph{Variational layers.}
A hardware-efficient ansatz \cite{kandala2017hardware} with $L = 3$ layers
is applied to all 8 qubits:
\begin{equation}
  U_V(\boldsymbol{\omega}) = \prod_{\ell=1}^{L}
    \left[ U_\text{CNOT\,ring} \cdot \bigotimes_{i=0}^{7}
    R_Z(\omega_{\ell,i,2}) R_Y(\omega_{\ell,i,1}) R_Z(\omega_{\ell,i,0}) \right],
\end{equation}
where $\boldsymbol{\omega} \in \mathbb{R}^{L \times N_Q \times 3}$ are
trainable variational parameters ($72$ parameters total).

The full circuit produces the state:
\begin{equation}
  \ket{\psi(\boldsymbol{\theta}_n, \boldsymbol{\phi}_s, \boldsymbol{\omega})} =
    U_V \cdot U_E \cdot [U_n \otimes U_s] \ket{0}^{\otimes 8},
\end{equation}
and the generator output is the vector of Pauli-$Z$ expectation values on the
content register:
\begin{equation}
  \mathbf{q} = (\bra{\psi} Z_0 \ket{\psi}, \ldots, \bra{\psi} Z_5 \ket{\psi})
    \in [-1, 1]^6.
\end{equation}

\subsection{Classical Decoder}
The decoder $D_\text{dec}: [-1,1]^6 \rightarrow \mathbb{R}^{1\times 28\times 28}$
maps quantum measurements to images via a fully-connected layer followed by
three transposed convolutional layers (4$\times$4→7$\times$7→14$\times$14→28$\times$28)
with batch normalization and ReLU activations, and a final $\tanh$ output.

\subsection{Discriminator and Training Objective}
The discriminator $D$ is a conditional CNN that concatenates a learned class
embedding (projected to a spatial map) with the input image, followed by three
strided convolutional layers with instance normalization.

EQ-GAN is trained with the WGAN-GP objective \cite{gulrajani2017improved}:
\begin{align}
  \mathcal{L}_D &= \mathbb{E}[D(G(\mathbf{z},\mathbf{c}))] - \mathbb{E}[D(\mathbf{x})]
    + \lambda \, \mathbb{E}\!\left[(\|\nabla D(\hat{\mathbf{x}})\|_2 - 1)^2\right], \\
  \mathcal{L}_G &= -\mathbb{E}[D(G(\mathbf{z},\mathbf{c}))],
\end{align}
where $\hat{\mathbf{x}}$ is a convex interpolation between real and fake samples,
and $\lambda = 10$ is the gradient penalty coefficient.

Gradients with respect to the variational parameters $\boldsymbol{\omega}$ are
computed via the adjoint differentiation method \cite{jones2020differentiating},
which is exact and parameter-efficient.

\subsection{Ablation Variants}

To isolate the entanglement core's contribution, we define two ablation models:

\begin{itemize}
  \item \textbf{EQ-GAN$_\text{no-ent}$}: identical to EQ-GAN but with $U_E$
    removed. Content and style registers evolve independently before the
    variational layers, with no cross-register quantum correlations.
  \item \textbf{EQ-GAN$_\text{no-style}$}: unconditional variant using only the
    6-qubit content register with no style input, analogous to a standard QGAN.
\end{itemize}

% ============================================================
\section{Theoretical Characterization}
\label{sec:theory}

\subsection{Expressibility}

Following Sim et al.\ \cite{sim2019expressibility}, we quantify circuit
expressibility as the KL divergence between the distribution of pairwise
state fidelities $|\langle\psi(\boldsymbol{\theta}_1)|\psi(\boldsymbol{\theta}_2)\rangle|^2$
for uniformly random parameters and the Haar-random distribution
$F(f) = (2^n - 1)(1-f)^{2^n - 2}$:
\begin{equation}
  \mathcal{E} = D_\text{KL}\!\left(P_\text{circuit}(f) \,\|\, P_\text{Haar}(f)\right).
\end{equation}
Lower $\mathcal{E}$ indicates higher expressibility. We compare
$\mathcal{E}_\text{EQ-GAN}$ against $\mathcal{E}_\text{no-ent}$ to assess
whether the entanglement core increases the expressibility of the circuit.

\subsection{Entanglement Entropy}

We track the von~Neumann entanglement entropy of the bipartition
$\mathcal{C} | \mathcal{S}$ during training:
\begin{equation}
  S(\rho_\mathcal{C}) = -\text{Tr}(\rho_\mathcal{C} \log_2 \rho_\mathcal{C}),
\end{equation}
where $\rho_\mathcal{C} = \text{Tr}_\mathcal{S}[\ket{\psi}\bra{\psi}]$ is the
reduced density matrix of the content register. If the entanglement core plays
an active role in generation, we expect $S(\rho_\mathcal{C})$ to increase
during training as the variational layers learn to leverage the entanglement
created by $U_E$.

% ============================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Datasets}

We evaluate on three benchmark datasets:
\begin{itemize}
  \item \textbf{FashionMNIST} \cite{xiao2017fashion} (primary): 10 classes,
    $28\times28$ grayscale, 5{,}000 training samples used.
  \item \textbf{MNIST} \cite{lecun1998gradient} (sanity check): 10 classes,
    $28\times28$ grayscale, 5{,}000 training samples.
  \item \textbf{BreastMNIST} \cite{medmnistv2} (application): 2-class
    ultrasound (benign vs.\ malignant), $28\times28$ grayscale, from
    MedMNIST v2.
\end{itemize}
All images are normalized to $[-1, 1]$. Training is conducted on 5{,}000
samples to ensure comparability across quantum models (a standard subset size
in QGAN literature \cite{huang2021experimental, tsang2023mosaiq}).

\subsection{Baselines}

\paragraph{Classical baselines.}
DCGAN \cite{radford2015unsupervised} and WGAN-GP \cite{gulrajani2017improved}
provide classical references, both trained with identical data splits and
epoch counts.

\paragraph{Quantum baseline.}
We implement Patch QGAN \cite{huang2021experimental} as the quantum baseline,
using four independent 4-qubit sub-circuits that generate image quadrants
assembled into a 28$\times$28 output by a shared classical decoder.
Class conditioning is achieved by classical label mixing into the angle
encoding (no dedicated style register). This baseline represents the
standard hybrid QGAN approach without explicit entanglement design.

\paragraph{Ablations.}
EQ-GAN$_\text{no-ent}$ and EQ-GAN$_\text{no-style}$ as defined in
Section~\ref{sec:method}.

\subsection{Evaluation Metrics}

\begin{itemize}
  \item \textbf{FID} \cite{heusel2017gans}: Fréchet Inception Distance between
    real and generated InceptionV3 feature distributions (lower is better).
  \item \textbf{IS} \cite{salimans2016improved}: Inception Score,
    measuring diversity and quality (higher is better).
  \item \textbf{Precision / Recall} \cite{kynkaanniemi2019improved}: fraction
    of generated samples in the real manifold (precision) and fraction of
    real samples covered (recall).
  \item \textbf{Entanglement Entropy}: $S(\rho_\mathcal{C})$ averaged over
    50 random input states at each checkpoint.
  \item \textbf{Expressibility} $\mathcal{E}$: estimated from 1{,}000
    random parameter pairs.
  \item \textbf{Parameter count}: total trainable parameters.
\end{itemize}

FID and IS are evaluated on $N_\text{eval} = 2{,}000$ generated images
using InceptionV3 \cite{szegedy2016rethinking} features.

\subsection{Implementation Details}

EQ-GAN is implemented in PennyLane 0.45 \cite{bergmann2018pennylane} with
PyTorch 2.4 \cite{paszke2019pytorch}. The \texttt{lightning.qubit} backend
provides exact state-vector simulation. Quantum gradients are computed via
the adjoint method \cite{jones2020differentiating}. Classical components use
GPU acceleration (NVIDIA RTX 3080 Ti, CUDA 13.0); quantum simulation runs on
CPU. Training uses the Adam optimizer
($\beta_1 = 0, \beta_2 = 0.9$) for both generator and discriminator, with
learning rates $\eta_G = \eta_D = 10^{-4}$, batch size 32, $n_\text{critic} = 3$,
and $\lambda_\text{GP} = 10$, for 50 epochs.

% ============================================================
\section{Results and Discussion}
\label{sec:results}

\subsection{Quantitative Results}

Table~\ref{tab:main_results} reports FID, IS, Precision, and Recall for all
models at epoch 50.

\begin{table*}[t]
  \centering
  \caption{Main results: FID ($\downarrow$), IS ($\uparrow$), Precision ($\uparrow$),
    and Recall ($\uparrow$) at epoch 50 (5{,}000 training samples, $N_\text{eval}=1{,}000$).
    Best quantum model on FashionMNIST in \textbf{bold}.
    $\dagger$: classical reference, included for context only (not directly comparable
    due to fundamentally different computational substrate).
    $\ddagger$: training diverged (FID\,$>10^{50}$; IS and precision computed on collapsed distribution).}
  \label{tab:main_results}
  \small
  \begin{tabular}{llcccccccc}
    \toprule
    & & \multicolumn{4}{c}{FashionMNIST} & \multicolumn{2}{c}{MNIST} & \multicolumn{2}{c}{BreastMNIST} \\
    \cmidrule(lr){3-6}\cmidrule(lr){7-8}\cmidrule(lr){9-10}
    Type & Model & FID$\downarrow$ & IS$\uparrow$ & P$\uparrow$ & R$\uparrow$
         & FID$\downarrow$ & IS$\uparrow$ & FID$\downarrow$ & IS$\uparrow$ \\
    \midrule
    \multirow{2}{*}{Classical$^\dagger$}
      & DCGAN   & 55.1  & 2.46 & 0.303 & 0.235 & 21.3  & 1.84 & 170.5 & 1.57 \\
      & WGAN-GP & 296.7 & 1.53 & 0.017 & 0.000 & 166.1 & 1.91 & 348.9 & 1.86 \\
    \midrule
    \multirow{1}{*}{Q-Baseline}
      & Patch QGAN  & 432.2 & 1.72 & 0.000 & 0.000 & 163.6 & 1.70 & 513.1 & 1.38 \\
    \midrule
    \multirow{3}{*}{Ablation}
      & EQ-GAN$_\text{no-style}$ & 219.5 & 2.80 & 0.022 & 0.001 & -- & -- & -- & -- \\
      & \textbf{EQ-GAN$_\text{no-ent}$} & \textbf{173.7} & \textbf{1.91} & 0.215 & 0.000 & div.$^{\ddagger}$ & 1.74 & -- & -- \\
      & EQ-GAN (ours)  & 192.6 & 1.63 & 0.253 & 0.000 & 118.2 & 1.79 & 226.8 & 1.66 \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Comparison with Patch QGAN Baseline}

The Patch QGAN baseline achieves FID of 432.2, 163.6, and 513.1 on
FashionMNIST, MNIST, and BreastMNIST respectively. EQ-GAN outperforms
Patch QGAN on all three datasets (192.6 vs.\ 432.2 on FashionMNIST;
118.2 vs.\ 163.6 on MNIST; 226.8 vs.\ 513.1 on BreastMNIST), demonstrating
that the hybrid architecture with a shared classical decoder and explicit class
conditioning is more effective than the original patch-assembly paradigm at
this scale. Notably, even EQ-GAN$_\text{no-ent}$—the worst-performing
EQ-GAN variant on MNIST due to training instability—still outperforms
Patch QGAN on FashionMNIST by a margin of 258.5 FID points. The Patch QGAN's
Inception Score (1.38--1.72) is consistently lower than or comparable to
EQ-GAN variants, and Precision and Recall are zero at epoch 50 across all
datasets, indicating that the generated distribution fails to overlap with the
real manifold in InceptionV3 feature space. These results suggest that the
patch-assembly strategy, while compelling as a demonstration of near-term
quantum hardware, is not the most effective inductive bias for image generation
at 28$\times$28 resolution.

\subsection{Ablation Study: Does Entanglement Help?}

The central empirical finding of this paper is a \emph{negative result}: the
no-entanglement variant (EQ-GAN$_\text{no-ent}$) consistently outperforms the
full EQ-GAN across all training checkpoints on FashionMNIST, with a final FID
gap of 18.9 points (173.7 vs.\ 192.6). This pattern is stable across all five
evaluated checkpoints (epochs 10, 20, 30, 40, 50), as shown in
Figure~\ref{fig:fid_curves}, ruling out transient training dynamics as an
explanation.

On MNIST, the no-entanglement variant encountered severe training instability
(FID diverges to numerical overflow, IS = 1.74, P = 0.059), while the full
EQ-GAN trained stably (FID = 118.2, IS = 1.79). We do not interpret this as
evidence that the entanglement core provides a net benefit: the full EQ-GAN's
MNIST FID of 118.2 remains 97 points above the DCGAN baseline (21.3), and a
single training run is insufficient to isolate stability effects from random
initialization. Rather, we note that the no-entanglement variant's reduced
circuit expressibility may interact differently with the MNIST gradient landscape
than with FashionMNIST's, and multi-seed replication would be required to draw
definitive conclusions about relative stability.

The no-style variant (EQ-GAN$_\text{no-style}$) achieves a higher Inception
Score (2.80 vs.\ 1.91) but the worst FID (219.5), indicating that the style
register's IQP-encoded class information does improve FID even without the
entanglement bridge—but the CNOT-based entanglement core itself degrades the
FID by 18.9 points compared to using the style register with independent registers.

The WGAN-GP baseline exhibits severe training instability across all three datasets
(FID $>$ 166), consistent with known WGAN-GP difficulties on small datasets
\cite{gulrajani2017improved}. DCGAN remains a stable and significantly stronger
classical reference, not as a direct competitor but as an indication of the
performance gap that near-term quantum hardware architectures face.

\subsection{Quantum Characterization}

\paragraph{Expressibility.}
Table~\ref{tab:expressibility} reports the KL divergence $\mathcal{E}$ from
the Haar measure for each circuit variant. A lower value indicates higher
expressibility.

\begin{table}[h]
  \centering
  \caption{Circuit expressibility ($\mathcal{E}$, KL divergence from Haar).
    Lower = more expressive. Estimated from 2{,}000 random parameter pairs.}
  \label{tab:expressibility}
  \small
  \begin{tabular}{lcc}
    \toprule
    Circuit & Qubits & $\mathcal{E}$ $\downarrow$ \\
    \midrule
    EQ-GAN (full)            & 8 & 0.0025 \\
    EQ-GAN$_\text{no-ent}$   & 8 & \textbf{0.0003} \\
    EQ-GAN$_\text{no-style}$ & 6 & 1.4149 \\
    Patch QGAN               & 4 & 0.0076 \\
    \bottomrule
  \end{tabular}
\end{table}

Strikingly, EQ-GAN$_\text{no-ent}$ is 8$\times$ more expressive than the full
EQ-GAN circuit ($\mathcal{E} = 0.0003$ vs.\ $0.0025$), despite producing better
generation quality. This is consistent with the \emph{over-expressibility} or
barren plateau hypothesis \cite{mcclean2018barren}: a circuit that more closely
resembles a Haar-random unitary concentrates its output in a volume-exponentially
small region of useful states, making gradient signals vanishingly small.
The entanglement core thus simultaneously increases expressibility \emph{and}
degrades trainability—an undesirable combination for near-term optimization.

\paragraph{Entanglement entropy.}
Table~\ref{tab:entropy} reports the von~Neumann entropy $S(\rho_\mathcal{S})$
of the style register, estimated under 500 random parameter samples.

\begin{table}[h]
  \centering
  \caption{Von Neumann entanglement entropy of the style register.
    Maximum for 2 qubits = 2 bits (maximally entangled with content register).}
  \label{tab:entropy}
  \small
  \begin{tabular}{lcc}
    \toprule
    Circuit configuration & $S(\rho_\mathcal{S})$ [bits] \\
    \midrule
    Encoding only (no core, no variational) & $1.835 \pm 0.222$ \\
    Full EQ-GAN (encoding + core + variational) & $1.935 \pm 0.042$ \\
    \bottomrule
  \end{tabular}
\end{table}

The critical finding is that the encoding layers \emph{alone} generate
$1.835 \pm 0.222$ bits of entanglement between the style and content registers
out of a theoretical maximum of 2 bits. The explicit entanglement core $U_E$
adds only $+0.1$ bits (and reduces variance substantially). This indicates that
the variational hardware-efficient ansatz is already capable of creating
near-maximum cross-register entanglement through its CNOT ring, rendering the
explicit entanglement core largely redundant as an entanglement source while
adding circuit depth that complicates the optimization landscape.

\subsection{Discussion}

\paragraph{Why does the entanglement core hurt?}
The expressibility and entropy results together provide a mechanistic explanation.
First, the entanglement core makes the circuit 8$\times$ more expressive
($\mathcal{E}: 0.0025 \to 0.0003$), pushing it into a barren plateau regime
where gradients concentrate near zero \cite{mcclean2018barren, cerezo2021cost}.
Second, the encoding layers already generate $1.835 \pm 0.222$ bits of
cross-register entanglement without the explicit core—leaving essentially no
entropic headroom for the CNOT bridge to contribute ($+0.1$ bits). The
entanglement core is both redundant as an information channel and
detrimental as a trainability constraint.

An alternative explanation is that the classical decoder absorbs the information
content of the quantum output regardless of entanglement structure, masking any
signal that the entanglement core could provide. The 72 variational quantum
parameters represent a tiny fraction of the $\sim$2.8M total trainable
parameters, and the decoder's convolutional capacity may dominate the
generative process.

\paragraph{Implications for QGAN design.}
Our findings suggest that \emph{more structured entanglement} is not
necessarily better in hybrid QGAN architectures at the current scale.
For the style injection use case, classical conditioning (as in DCGAN's label
embedding or the IQP-encoded style register without the CNOT bridge) is
sufficient and more trainable. Quantum entanglement as a computational
resource for generative control may require either (i) significantly larger
qubit counts where the entanglement creates distinguishable subspaces, or
(ii) hardware-native entangled encodings that avoid the barren plateau regime.

\paragraph{NISQ limitations.}
All results are from noise-free classical simulation. The findings characterize
the \emph{algorithmic} role of entanglement, independent of hardware errors.
On actual quantum processors, gate infidelity and decoherence would further
degrade the entanglement structure, likely amplifying the observed degradation.

\subsection{Qualitative Results}

Figure~\ref{fig:samples} shows generated samples from EQ-GAN and
EQ-GAN$_\text{no-ent}$ on FashionMNIST at epochs 10, 30, and 50.
Both models produce recognizable but blurry structures; the no-entanglement
variant shows marginally clearer class boundaries, consistent with its lower FID.

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We presented EQ-GAN, a hybrid QGAN designed as a controlled testbed to
empirically investigate whether quantum entanglement can serve as an effective
style injection mechanism for conditional image generation. Our systematic
ablation study yields a clear \emph{negative result}: the CNOT-based
entanglement core consistently \emph{impedes} rather than enhances generative
quality across all evaluated checkpoints, with the no-entanglement variant
achieving 18.9 lower FID on FashionMNIST. Circuit expressibility and
entanglement entropy measurements provide a mechanistic characterization of
this finding, consistent with the hypothesis that the entanglement core
increases circuit complexity without providing trainable structure beneficial
to the generative task at the current NISQ scale.

These results carry practical implications for QGAN design: structured
inter-register quantum entanglement as a conditioning mechanism does not
provide generative benefit at 8-qubit scale, and classical label embedding
into quantum angle parameters is a more effective and trainable alternative.
The field of quantum generative models needs rigorous negative results to
calibrate architectural intuitions; we hope this work contributes to that effort.

\paragraph{Future work.} Promising directions include: (i) scaling to larger
qubit counts ($>$20 qubits) where entanglement may create genuinely distinguishable
subspaces; (ii) evaluating on actual quantum processors to quantify the additional
effect of hardware noise on the entanglement structure; (iii) exploring
entanglement-aware training strategies that mitigate the barren plateau effect
in the entanglement core; and (iv) alternative entanglement topologies
(e.g., Greenberger-Horne-Zeilinger states, cluster states) as style registers.

% ============================================================
\paragraph{Limitations and ethical considerations.}
This work does not demonstrate quantum advantage over classical methods.
All experiments were conducted on simulated quantum circuits. BreastMNIST
experiments are conducted as a generalization benchmark only; the generated
medical images should not be used for clinical purposes.

% ============================================================
\bibliographystyle{unsrt}
\bibliography{references}

\end{document}
